
\documentclass[conference]{IEEEtran}

\usepackage[T1]{fontenc}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{url}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=black,
  citecolor=black
}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

\algrenewcommand\algorithmicrequire{\textbf{Input:}}
\algrenewcommand\algorithmicensure{\textbf{Output:}}

\title{Branch Prediction Simplification for AI Workloads:\\Progress Report}

\author{
\IEEEauthorblockN{Simon Darota, Winner Abula}
\IEEEauthorblockA{
Department of Computer Science\
Khalifa University\
Abu Dhabi, UAE\
\{100067950, 100066896\}@ku.ac.ae
}
}


\begin{document}
\maketitle

\begin{abstract}

State-of-the-art branch predictors usually achieve high accuracy by combining multiple history lengths, tags, specialized update logic, and auxiliary correction techniques. This kind of design mechanism can be justified for general purpose complex workloads, however, it may not be necessary for AI workloads whose control flow is usually filled with regular iterative spaces.
In this project, we study whether a simpler loop-aware branch predictor is able to support competitive performance on AI based workloads simultaneously reducing predictor complexity and storage. To prove this point, we created a gem5-based evaluation pipeline for RISC-V O3CPU and compared low-complexity baseline predictors (BiModeBP and GshareBP) with advanced high-complexity predictor (TAGE). We also built a new Loop-Aware Branch Predictor (LABP) specifically tailored to AI workloads and compared its performance with the three baseline branch predictors.  Preliminary results on an inference-style microbenchmark task (\texttt{tinymlp}) demonstrates that TAGE remains far superior in terms of both Instructions Per Cycle (IPC) and misprediction rate, but our LABP underperforms the simpler low-complexity baselines. This result shows loop-awareness alone is insufficient unless arbitration andconfidence mechanisms are carefully designed. This report defines the problem, explains and critiques existing approaches, presents the evaluation pipeline along with preliminary quantitative findings. 
\end{abstract}

\section{Problem Description}
In computer architecture, branch predictor is a digital circuit that attempts to guess which way a branch (for example, if-then-else statement) will go prior to this is known for sure. A wrong-path fetch decision wastes front-end bandwidth, pollutes speculation and reduces effective instruction throughput, hence, conditional branch prediction remains critical to high-performance processors. In deep and wide out-of-order pipelines in which a single misprediction can make substantial speculative work obsolete, the penalty is especially severe. For this reason, modern branch predictors are no longer just small counter tables, they are sophisticated multi-component designs with long histories and explicit-anti-aliasing techniques~\cite{smith_1981,yeh_patt_1992,mcfarling_1993,seznec_michaud_2006}.

This evolution creates a lot of tension. High-complexity predictors like TAGE and its derivatives achieve very high accuracy across variety of workloads, however, they also require multiple tagged tables, allocation policies that are nontrivial, heuristics that are update and extra storage ~\cite{seznec_michaud_2006,seznec_2016}. That complexity is usually justified for a general-purpose complex core. The answer is less obvious for AI based workloads. The kernels in dense linear regression models or deep neural-network inference are put together as regular loop nests. A simpler predictor may recover most of the achievable prediction performance at lower implementation cost if their dynamic branch stream is dominated by highly biased loop branches. 

The central question of this project is therefore:
\begin{quote}
\emph{Can a lightweight, loop-aware branch predictor achieve competitive accuracy and performance on AI workloads relative to conventional predictors such as bimodal, gshare, and TAGE, while reducing predictor complexity and conditional predictor storage?}
\end{quote}

This question matters for two main reasons. First, branch prediction is an energy-sensitive structure because it is consulted every cycle on the fetch path. Second, AI-oriented processors and energy-constrained systems benefit disproportionately from simpler structures when performance loss is small. Our hypothesis does not invalidate the necessity of complex predictors totally; it rather emphasizes the fact that the control-flow structure of some AI workloads may justify the need for a more specialized design.

This project contract motivated a loop-aware predictor for exactly this scheme. The intended target is a workload in which loop-exit branches are common, strongly patterned, and potentially predictable with much less machinery than a general-purpose tagged predictor. The present progress report evaluates that hypothesis through an initial gem5 implementation and a first benchmark study.

\section{Related Works}
Branch prediction research offers a spectrum of design points, from very small tables to highly specialized hybrid mechanisms. This section reviews the most relevant approaches and identifies what each does well, where it breaks down, and why it matters for AI-oriented control flow.

\subsection{Bimodal Predictors}
The bimodal predictor is the canonical low-complexity baseline: a table of saturating counters indexed by branch PC bits~\cite{smith_1981}. Its strengths include being compact, fast, straightforward to verify, and effective on strongly biased branches. When simplicity is our design goal, these properties make it attractive.

It is also not without a weakness. Bimodal prediction cannot exploit inter-branch correlation and is highly vulnerable to aliasing. Once multiple branches map to the same entry, unrelated behavior interferes destructively. As a result, bimodal designs generally perform poorly on branches whose outcomes depend on recent path history rather than fixed bias. 

For this project, bimodality is valuable as a floor: if a proposed simplified design cannot consistently outperform a small bimodal table on the target workload class, the simplification argument is weak.

\subsection{Two-Level Adaptive Predictors and Gshare}
Two-level adaptive predictors extended the field by introducing explicit branch history into the indexing function~\cite{yeh_patt_1992}. Gshare is the classic practical instance: it XORs global history with PC bits to reduce destructive interference while preserving low implementation overhead~\cite{mcfarling_1993}. Compared to bimodal, gshare captures correlation among branches, and it usually provides better accuracy with trade off on mixed workloads.

The limitation is that its effectiveness depends on whether useful correlations fit within the available history and table budget. When history is too short, difficult branches remain unresolved. In addition to that, when the budget is small, aliasing remains substantial. More importantly for AI workloads, gshare is agnostic to loop semantics. It benefits from repeated patterns, but it does not explicitly represent iteration count or loop confidence. Thus, it can still mispredict loop exits that are regular in a semantic sense but difficult to encode in a short global-history pattern.

\subsection{Hybrid and Interference-Reducing Predictors}
Hybrid prediction emerged to address the fact that no single mechanism works best for all branches. Tournament-style combinations and related designs select among multiple predictors so that different components specialize on different branch classes~\cite{mcfarling_1993}. Work on correlated branch prediction also highlighted how interference, rather than pure table size, often limits accuracy~\cite{young_1995}. These designs are critical as they show that moderate increment in complexity can bring substantial gains over a single predictor table.

Their drawback, from the perspective of this project, is that they move the design away from the simplicity notion. Hybrid predictors require extra tables, a selector, and more involved update logic. They often represent an effective engineering compromise, but they do not answer the narrower question posed here: whether a predictor can be simplified specifically for loop-heavy AI kernels rather than made more general.

\subsection{Tagged Predictors}
Tagged predictors deal with aliasing more directly. Michaud's tag-based work and the later TAGE design established a highly influential design principle: maintain predictor components indexed with different history lengths and use partial tags so the predictor can match the right context while limiting destructive interference~\cite{michaud_2005,seznec_michaud_2006}. TAGE is especially effective because it can use short history for easy branches and long history for difficult ones, with tagging providing robustness that simple two-level predictors lack.

For our purposes, TAGE is the strongest reference design. Its primary advantage is breadth: it delivers high accuracy across diverse branch behaviors, including branches that require distinct correlation depths. The limitation is complexity. TAGE requires multiple tagged tables, usefulness tracking, allocation and replacement logic, and a careful update policy. Since our study is centered on simplicity, TAGE will be our performance scale instead of implementation point.

\subsection{Loop Predictors}
Loop predictors exploit a specific regularity that other predictors only learn indirectly: the repeated not-taken/taken structure of counted loops. By tracking loop trip count, confidence, and iteration progress, a loop predictor can remove the persistent loop-exit mispredictions that remain even under strong global-history schemes~\cite{seznec_2016}. This is the mechanism that motivates this project.

What is good about loop prediction is also its main limitation. It is excellent when the branch stream actually contains stable counted loops, but is narrow in scope. A loop predictor can only help when it detects the right structure, maintains confidence correctly, and overrides the base predictor conservatively. On branches driven by input sparsity, activation patterns, or other data-dependent conditions, loop-awareness provides little value and can become harmful if override logic is too aggressive. Thus, loop prediction is promising for AI workloads only if those workloads are truly dominated by regular loop control rather than merely containing loops.

\subsection{Perceptron and Neural Predictors}
Perceptron-based branch prediction extended the field in a different direction by learning long-range correlations using weight vectors and dot products~\cite{jimenez_lin_2001,jimenez_lin_2002}. These methods can outperform table-based predictors based on shorter histories in branches with complex dependence on distant history.

For this project, neural predictors serve mainly as a contrast case. They reinforce a broader lesson from the literature: strong accuracy on arbitrary workloads usually requires more machinery, not less. Although perceptron predictors are conceptually elegant, they increase hardware complexity and are not aligned with the project's simplification objective. Their existence strengthens the relevance of our question by showing how far the field has moved toward complexity in pursuit of generality.

\subsection{Recent Perspective and Relevance to AI Workloads}
Recent work continues to emphasize the trade-off between predictor accuracy, storage efficiency, and implementation cost in contemporary RISC-V settings~\cite{sutar_guinde_2025}. At the same time, workload characterization studies show that branch predictability varies significantly across programs and inputs; the branch stream cannot be inferred from application labels alone~\cite{vikas_gratz_jimenez_2025}. This is the key caution for AI-oriented prediction. Dense kernels may be dominated by regular loops, but sparsity, conditional execution, and data-dependent pruning can introduce branches that are neither highly biased nor easily modeled by loop trip counting. A simplified predictor is therefore plausible only for a subset of AI workloads, not by default for all of them.

\section{Proposed Solution}
The proposed design is a \textbf{Loop-Aware Branch Predictor (LABP)} intended to occupy a narrower but simpler design point than TAGE. The architecture has three components:

\begin{enumerate}
    \item \textbf{Base predictor:} a small PC-indexed bimodal table of 2-bit saturating counters.
    \item \textbf{Loop predictor:} a compact loop structure that tracks candidate loop branches, iteration count, and confidence.
    \item \textbf{Override policy:} the loop predictor may override the base prediction only when it has a valid and sufficiently confident loop prediction.
\end{enumerate}

The underlying hypothesis is targeted and specific rather than universal. If the dominant hard branches in an AI kernel are loop exits, then explicit loop tracking should remove many mispredictions. This would otherwise require a larger general-purpose predictor. A bimodal base is retained because it is simple, fast, and adequate for strongly biased non-loop branches. The loop component is meant to add specialization without expanding into a full hybrid tagged design.

\subsection{Implementation Strategy in gem5}
LABP was implemented directly in gem5 as a new \texttt{ConditionalPredictor}. The predictor logic is defined in new source files under the branch-prediction subsystem, compiled through gem5's build system, and exposed through the Python configuration layer so that runs can be selected with \texttt{--bp-type=LAP}. The evaluation flow uses the same gem5 O3CPU configuration for all predictors in order to isolate branch-prediction effects. Metrics are extracted from \texttt{stats.txt}, while conditional predictor storage is estimated from generated \texttt{config.ini} parameters.

In our design we reuse gem5's loop-predictor support rather than re-implementing loop tracking from scratch. This keeps the proposed predictor modular and allows the study to focus on the composition of a simple base predictor with a loop-aware override. The main research question is therefore not whether loop prediction can be implemented, but whether this particular combination is sufficient to recover a useful fraction of TAGE's performance on the target workload class.

\subsection{Predictor Pseudocode}
Algorithm~\ref{alg:lapbp} summarizes the predictor behavior at a higher level. The essential point is that LABP computes a low-cost base prediction first, then admits a loop-based override only when the loop component is confident.

\begin{algorithm}[t]
\caption{LABP lookup and update flow}
\label{alg:lapbp}
\footnotesize
\begin{algorithmic}[1]
\Require PC, thread id $tid$, actual outcome on update
\State \textbf{State:} bimodal counter table $B[0 \ldots N-1]$, loop table $L$
\Function{Lookup}{$tid, pc$}
    \State $idx \gets (pc \gg shift) \;\&\; (N-1)$
    \State $basePred \gets (B[idx] > \theta)$
    \State $(loopValid, loopPred, meta) \gets L.\textsc{Predict}(tid, pc)$
    \If{$loopValid$}
        \State $finalPred \gets loopPred$
    \Else
        \State $finalPred \gets basePred$
    \EndIf
    \State \Return $(finalPred,\{idx, basePred, meta\})$
\EndFunction
\Function{Update}{$tid, pc, actual, h, squashed$}
    \If{$h = \varnothing$}
        \State \Return
    \EndIf
    \If{$squashed$}
        \State $L.\textsc{Squash}(h.meta)$
        \State \Return
    \EndIf
    \State $L.\textsc{Update}(tid, pc, actual, h.meta, h.basePred)$
    \If{$actual = taken$}
        \State Increment $B[h.idx]$
    \Else
        \State Decrement $B[h.idx]$
    \EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Benchmark Structure}
The first benchmark is \texttt{tinymlp}, a compact inference-style microbenchmark. It contains the loop nests one would expect in an MLP like forward pass, but also introduces data-dependent conditions that skip work when values are zero or below a sparsity threshold. The resulting control flow is therefore mixed: regular loop branches coexist with input-dependent inner branches. This makes \texttt{tinymlp} a useful first test because it probes exactly the boundary case that matters for our project: a workload that appears loop-heavy at a structural level but is not purely regular at the branch level.

\begin{algorithm}[t]
\caption{\texttt{tinymlp} control-flow skeleton}
\label{alg:tinymlp}
\footnotesize
\begin{algorithmic}[1]
\For{$iter = 1$ to $T$}
    \For{$i = 1$ to $H$}
        \State $acc \gets b_1[i]$
        \For{$j = 1$ to $D$}
            \If{$|x[j]| > \tau$}
                \State $acc \gets acc + W_1[i,j] \cdot x[j]$
            \Else
                \State $skipped \gets skipped + 1$
            \EndIf
        \EndFor
        \State $h[i] \gets \mathrm{ReLU}(acc)$
    \EndFor
    \For{$k = 1$ to $O$}
        \State $acc \gets b_2[k]$
        \For{$i = 1$ to $H$}
            \If{$h[i] \neq 0$}
                \State $acc \gets acc + W_2[k,i] \cdot h[i]$
            \Else
                \State $skipped \gets skipped + 1$
            \EndIf
        \EndFor
        \State $y[k] \gets acc$
    \EndFor
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Preliminary Results}
The current evaluation uses gem5 in RISC-V SE mode with O3CPU and caches enabled. All predictors were run on the same \texttt{tinymlp} binary for 10 million committed instructions. The analysis pipeline extracts IPC, CPI, branch predictor lookups, branch mispredictions, and an estimate of conditional predictor storage from gem5's output files. Table~\ref{tab:prelim} reports the current results.

\begin{table}[t]
\centering
\caption{Preliminary results on \texttt{tinymlp} (RISC-V O3CPU, 10M instructions).}
\label{tab:prelim}
\begin{tabular}{lrrrr}
\toprule
Predictor & IPC & CPI & Mispred. Rate & Cond. BP KB \\
\midrule
BiModeBP & 1.919 & 0.521 & 0.0386 & 4.000 \\
GshareBP & 1.007 & 0.993 & 0.0440 & 0.625 \\
TAGE & 3.661 & 0.273 & 0.0018 & 6.688 \\
LABP & 0.569 & 1.758 & 0.1216 & 5.501 \\
\bottomrule
\end{tabular}
\end{table}

Here are the three insights from the current outputs.

First, TAGE remains the strongest design by a wide margin. Its misprediction rate is more than an order of magnitude lower than the simple baselines, and that advantage translates directly into the highest IPC. This is expected from the tagged multi-history literature: when branch behavior is heterogeneous, the ability to match multiple correlation lengths and suppress aliasing is decisive~\cite{michaud_2005,seznec_michaud_2006,seznec_2016}.

Second, the present LABP design does \emph{not} yet validate the simplification hypothesis. It is worse than both BiModeBP and GshareBP in IPC and in misprediction rate, despite using more conditional predictor storage than either simple baseline. The result indicates that simply attaching a loop predictor to a bimodal base does not automatically produce a good predictor for AI-like code.

Third, the failure mode is informative. The most plausible explanation is that \texttt{tinymlp} is not dominated by clean counted-loop exits; instead, many dynamic branches come from sparsity checks and activation-dependent conditions. Those branches are not naturally predicted by loop-iteration tracking. If the loop component overrides the base predictor too aggressively, or if loop confidence is not selective enough, the loop-aware structure can actively harm accuracy. In addition, a simplified loop-aware predictor may still be viable, but only for workloads whose branch stream is actually dominated by regular loop control and only when arbitration is conservative.

The present data therefore establish a clear research direction for the next stage of the project. LABP must be refined around confidence gating and evaluated on additional kernels that separate counted-loop behavior from data-dependent inner-loop branching. Only then can the project determine whether branch prediction simplification is a real opportunity for AI workloads or merely an attractive but workload-sensitive intuition.

\bibliographystyle{IEEEtran}
\begin{thebibliography}{12}

\bibitem{smith_1981}
J. E. Smith, ``A study of branch prediction strategies,'' in \emph{Proc. 8th Annu. Symp. Computer Architecture}, 1981, pp. 135--148.

\bibitem{yeh_patt_1992}
T.-Y. Yeh and Y. N. Patt, ``Alternative implementations of two-level adaptive branch prediction,'' in \emph{Proc. 19th Annu. Int. Symp. Computer Architecture}, 1992, pp. 124--134.

\bibitem{mcfarling_1993}
S. McFarling, ``Combining branch predictors,'' Digital Western Research Laboratory, Tech. Rep. TN-36, 1993.

\bibitem{young_1995}
C. C. Young, N. Gloy, and M. D. Smith, ``A comparative analysis of schemes for correlated branch prediction,'' in \emph{Proc. 22nd Annu. Int. Symp. Computer Architecture}, 1995, pp. 276--286.

\bibitem{michaud_2005}
P. Michaud, ``A PPM-like, tag-based predictor,'' \emph{Journal of Instruction-Level Parallelism}, vol. 7, 2005.

\bibitem{seznec_michaud_2006}
A. Seznec and P. Michaud, ``A case for (partially) TAgged GEometric history length branch prediction,'' \emph{Journal of Instruction-Level Parallelism}, vol. 8, 2006.

\bibitem{seznec_2016}
A. Seznec, ``TAGE-SC-L branch predictors again,'' in \emph{Proc. 5th JILP Workshop on Computer Architecture Competitions}, 2016.

\bibitem{jimenez_lin_2001}
D. A. Jim\'enez and C. Lin, ``Dynamic branch prediction with perceptrons,'' in \emph{Proc. 7th Int. Symp. High-Performance Computer Architecture}, 2001, pp. 197--206.

\bibitem{jimenez_lin_2002}
D. A. Jim\'enez and C. Lin, ``Neural methods for dynamic branch prediction,'' \emph{ACM Trans. Computer Systems}, vol. 20, no. 4, pp. 369--397, 2002.

\bibitem{sutar_guinde_2025}
D. G. Sutar and N. B. Guinde, ``Towards memory-efficient and high-performance branch prediction: the LXOR architecture for control flow optimization in embedded and general-purpose RISC-V processors,'' \emph{Journal of Low Power Electronics and Applications}, vol. 15, no. 4, p. 64, 2025.

\bibitem{vikas_gratz_jimenez_2025}
Vikas, P. Gratz, and D. A. Jim\'enez, ``Workload characterization for branch predictability,'' arXiv:2512.15827, 2025.

\bibitem{gem5_101}
gem5 Project, ``gem5 101,'' [Online]. Available: \url{https://www.gem5.org/documentation/learning_gem5/gem5_101/}

\end{thebibliography}

\end{document}